---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First, check if we have the activity.csv file.
If not, we unzip it from the activity.zip file.
Then we read it to the data frame.
```{r loadata}
if (!file.exists("activity.csv")) {
    unzip("activity.zip")
}
data <- read.csv("activity.csv")
```
Looks like there are days (for example the first day 2012-10-01) where none of the steps were measured so they are all NAs.
```{r}
head(data)
```



## What is mean total number of steps taken per day?
Examine the data per day, I look at the total number of steps, mean of steps and I count how many intervals have measurements.

```{r}
require(plyr)
data.per.day <- ddply(data,"date",summarise, 
                      sum=sum(steps,na.rm=TRUE),
                      mean=mean(steps,na.rm=TRUE), 
                      count=sum(!is.na(steps)))
head(data.per.day)

```
This is a histogram of the total number steps taken each day.
```{r}
hist(data.per.day$sum, xlab='Total number of steps per day', 
     main='Histogram of total number of steps')
```

And we can see the **mean** (9354) and **median** (10395) total number of setps taken per day.
```{r}
mean(data.per.day$sum)
median(data.per.day$sum)
# summary(data.per.day$sum)
```

## What is the average daily activity pattern?
Examine the data per interval (across all days)
```{r}
data.per.interval <- ddply(data,"interval",summarise,
                           sum=sum(steps,na.rm=TRUE),
                           mean=mean(steps,na.rm=TRUE))
head(data.per.interval)
```
And we can show a time series plot of the 5-minute intervals and the average number of steps taken.
```{r}
plot(data.per.interval$interval,data.per.interval$mean,type="l",
     xlab="Interval",ylab="Average number of steps")
```

We find the interval with maxinum average number of steps (across all days). This interval turns out to be 835 or the time 8:35 AM to 8:40 AM
```{r}
max <- max(data.per.interval$mean)
subset(data.per.interval,mean==max)
```

## Imputing missing values

We can find out that there are 2304 rows of NA values in the origianl dataset.
```{r}
sum(is.na(data$steps))
```

We will just fill in the missing values with the average (across all days) number of steps for that interval
```{r}
data2 <- ddply(data, .(interval), mutate, avg=mean(steps,na.rm=TRUE))
data2[is.na(data2$steps),]$steps <- data2[is.na(data2$steps),]$avg 
```

With the mising values filled in, I again look at the total number of steps and mean of steps.
```{r}
data2.per.day <- ddply(data2,"date",summarise, 
                      sum=sum(steps,na.rm=TRUE),
                      mean=mean(steps,na.rm=TRUE), 
                      count=sum(!is.na(steps)))
head(data2.per.day)
```

This is a histogram of the total number steps taken each day after the missing values were filed in.
```{r}
hist(data2.per.day$sum, xlab='Total number of steps per day (no NAs)', 
     main='Histogram of total number of steps (no NAs)')
```

With missing values filled in the **mean** and **median** total number of setps taken per day both became 10766 - the previous mean.  This is as expected.
```{r}
mean(data2.per.day$sum)
median(data2.per.day$sum)
# summary(data2.per.day$sum)
```

Introduce another column "datetype" which tells us if the date was a weekday or not.
```{r}
getDateType <- function(x) {
    if (weekdays(as.Date(x),abbreviate=TRUE) %in% c("Sat","Sun")) {
		"weekend"
	} else {
		"weekday"
	}
}
data3 <- data2
data3$datetype <- sapply(data3$date, getDateType) 
data3$datetype <- factor(data3$datetype)
head(data3,n=7)
```

Now I again examine the steps per interval (across all days), but differentiate if the date was a weekday or not.
```{r}
data3.per.interval <- ddply(data3,.(interval,datetype),summarise,
                           sum=sum(steps,na.rm=TRUE),
                           mean=mean(steps,na.rm=TRUE))
head(data3.per.interval)
```
We show a panel plot of average number of steps per interval.

```{r}
library(lattice)
xyplot( mean ~ interval | datetype, data = data3.per.interval, layout = c(1,2),
        type="l", ylab="Number of steps")
```

